From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations

### 2015-12-17 manuscript/write_up.tex
revamped simulation

\author{Robert O. Ness}
\email{nessr@purdue.edu}
\affiliation[Purdue University]{Purdue University Department of Statistics, West Lafayette}
\alsoaffiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
\author{Karen Sachs}
\affiliation[Stanford University]{School of Medicine, Stanford University, Palo Alto}
\author{Olga Vitek}
\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}

@@

\added{Modern high-throughput technologies such as mass spectrometry simultaneously quantify hundreds or even thousands of biomolecular analytes.  Statistical associations (e.g. Pearson correlation, Spearman correlation, and mutual information) between the analytes  can suggest an enticing number of hypotheses regarding the underlying causal biomolecular mechanism.  However, associations do not imply causation. Formal methods of causal inference \cite{pearl2009causality} \todo{Could you also reference http://www.biomedcentral.com/1471-2105/8/S6/S5} are required to probe these statistical associations for causality, and infer the underlying regulatory mechanisms. }

\added{There is an increasing interest in causal inference in proteome research. It has previously been used to learn the directed structure of signal transduction networks, e.g. in flow cytometry investigations in Sachs et al. 2005 \cite{sachs2013single}, antibody/bead-based XMAP technology from Luminex in Saez-Rodriguez et al. 2009\cite{saez2009discrete} and others \todo{what do you mean by 'others'? Other technologies? Other authors? What did they do?} \cite{prill2011crowdsourcing}.
However, the successful examples of causal inference in proteomics are very few \cite{bensimon2012mass}. In this perspective we demonstrate that this is not an accident, because large numbers of quantified analytes can make the task of causal inference more difficult.  The difficulty stems from the fact that large numbers of quantified analytes  lead to more spurious pairwise associations, obfuscate the true signal, and increase the false discoveries of putative causal events. Below we describe in non-technical terms the process of elucidating causal associations from high-throughput data, and suggest practical approaches for causal inference in large scale proteomic datasets.  Specifically, we suggest that the task of causal inference can be facilitated by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses. }

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=1\textwidth]{figs/mapk.png}}
\caption{A: Overview of the 3 steps of causal inference, illustrated for the MAPK signaling cascade.    B, C, and D feature an experiment simulated from the Huang-Ferrell computational model of the phosphorylation cascade.  B: Pairwise plots of concentration values of phosphorylated (doubly phosphorylated for Mek and Erk) forms of each protein, and observed Spearman correlation values.  The Raf -- Erk correlation is high, despite the fact that Raf does not directly regulate Erk.  C: Concentrations of Raf versus Erk, where samples corresponding to high Mek (here, set to the top quartile) are highlighted \added{with filled circles}.  The right panel of C shows the subset of samples with high Mek (i.e. conditional on Mek being high). In these samples the association between Raf and Erk disappears, \added{and we infer} that Raf and Erk are {\it conditionally independent} given Mek.  D: After Raf is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears. A second simulation, which is similar to the previous one but incorporates the inhibition of Mek reveals the causal flow from Raf to Mek to Erk. \todo{Where is this last simulation shown? If it's not shown, add (not shown) at the end of the sentence.}
\label{mapkInference}}
\end{figure}

Figure \ref{mapkInference}A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.  In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.  Step 2 reduces the correlation graph to a sparse graph of \added{inferred} underlying conditional dependencies (Raf--Mek, and Mek--Erk).  Step 3 interrogates this graph to find putative directions of causal relationships (Raf-->Mek, and Mek-->Erk).  While step 1 has little requirements, step 2 requires multiple samples or replicates, and step 3 requires systematic interventions (e.g. with protein inhibitors).

Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the protein concentrations across the biological samples, and Spearman correlations to quantify the extent of the associations.  The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.  The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.  What about the non-causal Raf--Erk edge? Despite the high Raf--Erk correlation, there is no direct causal mechanism between them (aside from the one via Mek, which is already accounted for via the Raf-->Mek and Mek-->Erk edges).  In causal  inference, our goal is to eliminate this ``nuisance" edge.  How is this done?

\added{To describe Step 2 of causal inference, we introduce some terminology. If concentrations of two proteins vary between the biological samples in a coordinated manner, such that knowing the behavior of one protein provides information on the other, the two proteins are called {\it dependent}. Otherwise they are called {\it independent}. {\it Conditional independence} is a special case that is important to causal inference.  Two {\it dependent} proteins are {\it conditionally independent} if, after knowing (in probability language, {\it conditioning on}) the behavior of third-party biomolecules, the two proteins become independent. In other words, knowing the concentration of one protein provides no information about the other after considering the information from the third party.}

\added{While statistical associations and correlations are properties of the observed data, dependence and conditional independence are properties of the underlying processes that generate the data. Step 2 relies on statistical inference\cite{pearl2009causality} to infer from the observed data pairs of proteins that are conditionally independent. The conditionally independent pairs are ignored, and the remaining pairs are kept as potential causal relations.}

Let's see how this applies to the MAPK signaling cascade. \added{Since Raf regulates the concentration of Erk, Raf and Erk are dependent.  However, since the concentration of Erk is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.  Thus Raf and Erk are conditionally independent given Mek.  }  Figure \ref{mapkInference}C illustrates the process of statistical inference by comparing the concentrations of Raf and Erk. When we subset the measurements to only the samples with high Mek, we can no longer see the association between Raf and Erk. Formally, the algorithms repeatedly test the null hypothesis of conditional independence between Raf and Erk given Mek, and makes the decision based on p-values\cite{spirtes2000causation}. Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.

\added{At Step 2 in the figure,} the direction of the regulation remains unknown. Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses. Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor. The intervention does not affect the concentration of Mek, however it blocks its ability to phosphorylate other proteins.  After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.  From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.  With the intervention, we can finally move from the undirected graph in panel A - step 2 to the causal graph in panel A - step 3.

\added{Causation and dependence are linked as follows;
\begin{itemize}
\item For two proteins, if one is the direct cause of the other, they are dependent, even after conditioning on other proteins in the system.
\item If the two proteins that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other proteins in the system.
\end{itemize}}

In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related proteins. Step 1 creates a dense network of pairwise associations. \removed{Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of conditional independence between pairs of proteins.} \added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.  See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings. Numerous implementations of these algorithms are available in statistical software, e.g. in the R package {\tt bnlearn} \cite{scutari2009learning}.

@@

A typical high-throughput experiment includes a small number of interventions, a small number of biological replicates, and quantifies a large number of analytes such as proteins. This creates challenges in each of the 3 steps of causal inference above.

@@

We illustrate this problem with a computer simulation, inspired by Fan et. al \cite{fan2014challenges} but translated to our context. First, we simulated an experiment that quantifies the concentrations of 20 proteins in 100 biological samples.  Second, we simulates another experiment where the number of proteins was increased to 500.  In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution. In other words, we do not expect any biologically meaningful associations in these data. We repeated each of these simulations 500 times. Figure \ref{spur_corr} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation. As can be seen, the experiment with 500 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlations.  This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=1\textwidth]{figs/spurious_corr.png}}
\caption{Simulated experiments quantifying 20 (or 500) unrelated proteins in 100 biological samples. The histograms display the highest Pearson correlation across any pair of proteins, calculated over 500 repetitions of the simulation. Increasing the number of proteins results in higher values of Pearson correlation without any biological justification.}
\label{spur_corr}
\end{figure}

Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.  If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}. To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.  Instead of finding the highest spurious correlation between pairs of proteins, we apply an example of type of causal inference algorithm that performs a series of conditional independence tests between the sets of proteins (constraint-based algorithms\cite{spirtes2000causation}), and count the number of pairs proteins \added{where the null hypothesis of conditional independence was rejected}.  As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, \added{no test should ever be rejected, and any instance of rejection by this algorithm would have biological justification}.  We again repeat these experiments 500 times. Figure \ref{spur_dep} shows the histograms of the counts of false positives.  The figure graphs the results of the Grow-Shrink algorithm \cite{margaritis2003learning}, \added{but we also repeated the simulation with the IAMB algorithm\cite{tsamardinos2003algorithms} and Inter-IAMB algorithm\cite{yaramakala2005speculative} and achieved similar results}.    The results demonstrate that an increase in the number of proteins leads to an increase in false positives. This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positives (rejecting the null hypothesis of conditional independence.}
\label{spur_dep}
\end{figure}

A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes. While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 relies on having many replicates relative to the number analytes.  Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high.  The quality of causal inference approaches decreases drastically as the numbers of replicates fall.

Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional independence.  Quantifying a large number of analytes introduces a  challenge for this step as well. As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.   We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships. \added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ proteins may be sufficient to infer causal events between $k$ analytes . Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.  Also, if the step 2 results in too many false positives for potential causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.

@@

\begin{enumerate}
\item \textit{Limit the number of analytes.} Even though a list of analytes quantifiable with high-throughput technologies grows larger, only use a subset of measurements that are both biologically relevant and technologically accurate. The length of the list is not as important as the quality of measurements on the key parts of the system.  If the broader biological system is well understood, it may be possible to design a targeted experiment that focuses on a specific network or pathway, and ask more specific questions of the data, such as the presence of a particular regulatory event.  The more specific the question, the less data are needed to make solid causal conclusions.

\item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.  This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.  \added{A review by Twerve and Saez-Rodriguez provides a contrast of technologies by their content and sampling throughput\cite{SaezRodriguez:2012kx}.}

\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}. See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset. \cite{sachs2005causal}.

>>>>

\author{Robert O. Ness}
\affiliation[Purdue University]{Purdue University Department of Statistics, West Lafayette}
\alsoaffiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}
\author{Karen Sachs}
\affiliation[Stanford University]{School of Medicine, Stanford University, Palo Alto}
\author{Olga Vitek}
\email{o.vitek@neu.edu}
\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Boston}

@@

\added{Modern high-throughput technologies such as mass spectrometry simultaneously quantify hundreds or even thousands of biomolecular analytes.  Statistical associations (e.g. Pearson correlation, Spearman correlation, and mutual information) between the analytes  can suggest an enticing number of hypotheses regarding the underlying causal biomolecular mechanism.  However, associations do not imply causation. Formal methods of causal inference \cite{pearl2009causality,markowetz2007inferring} are required to probe these statistical associations for causality, and infer the underlying regulatory mechanisms. }

\added{There is an increasing interest in causal inference in proteome research. It has previously been used to learn the directed structure of signal transduction networks, e.g. in flow cytometry investigations in Sachs et al. 2005 \cite{sachs2013single}, antibody/bead-based XMAP technology from Luminex in Saez-Rodriguez et al. 2009\cite{saez2009discrete} and participants in the DREAM4 Predictive Signaling Network Challenge\cite{prill2011crowdsourcing}.
However, the successful examples of causal inference in proteomics are very few \cite{bensimon2012mass}. In this perspective we demonstrate that this is not an accident.  The difficulty stems from the fact that large numbers of quantified analytes, combined with small sample size, lead to more spurious pairwise associations, obfuscate the true signal, and increase the false discoveries of putative causal events. Below we describe in non-technical terms the process of elucidating causal associations from high-throughput data, and suggest practical approaches for causal inference in large scale proteomic datasets.  Specifically, we suggest that the task of causal inference can be facilitated by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses. }

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.95\textwidth]{figs/mapk.png}}
\caption{A: Overview of the 3 steps of causal inference, illustrated for the MAPK signaling cascade.    B, C, and D feature an experiment simulated from the Huang-Ferrell computational model of the phosphorylation cascade.  B: Pairwise plots of concentration values of phosphorylated (doubly phosphorylated for Mek and Erk) forms of each protein, and observed Spearman correlation values.  The Raf -- Erk correlation is high, despite the fact that Raf does not directly regulate Erk.  C: Concentrations of Raf versus Erk, where samples corresponding to high Mek (here, set to the top quartile) are highlighted \added{with filled circles}.  The right panel of C shows the subset of samples with high Mek (i.e. conditional on Mek being high). In these samples the association between Raf and Erk disappears, \added{and we infer} that Raf and Erk are {\it conditionally independent} given Mek.  D:  After Mek is inhibited, the observed association between Raf and Mek remains, while the association between Mek and Erk disappears. This reveals the causal flow from Raf to Mek. 
\label{mapkInference}}
\end{figure}

Figure \ref{mapkInference}A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.  In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.  Step 2 reduces the correlation graph to a \added{sparse graph of inferred} underlying conditional dependencies (Raf--Mek, and Mek--Erk).  Step 3 interrogates this graph to find putative directions of causal relationships (Raf-->Mek, and Mek-->Erk).  While step 1 has little requirements, step 2 requires multiple samples or replicates, and step 3 requires systematic interventions (e.g. with protein inhibitors).

Figure \ref{mapkInference}B illustrates Step 1 of the causal inference, and shows 2-way plots of the protein concentrations across the biological samples, and Spearman correlations to quantify the extent of the associations. \added{We chose Spearman correlation instead of Pearson correlation because it best quantifies nonlinear relationships.}  The correlation values are high, and would meet most reasonable cut-off thresholds for constructing the correlation network in the left part of panel A.  The Raf--Mek and the Mek--Erk correlation edges match the Raf-->Mek, Mek-->Erk known causal edges.  What about the non-causal Raf--Erk edge? Despite the high Raf--Erk correlation, there is no direct causal mechanism between them (aside from the one via Mek, which is already accounted for via the Raf-->Mek and Mek-->Erk edges).  In causal  inference, our goal is to eliminate this ``nuisance" edge.  How is this done?

\added{To describe Step 2 of causal inference, we introduce some terminology. If concentrations of two proteins vary between the biological samples in a coordinated manner, such that knowing the concentration of one protein provides information on the concentration of the other, the two proteins are called {\it dependent}. Otherwise they are called {\it independent}. {\it Conditional independence} is a special case that is important to causal inference.  Two {\it dependent} proteins are {\it conditionally independent} if, after knowing (in probability language, {\it conditioning on}) the concentration of third-party biomolecules, the two proteins become independent. In other words, after considering the information from the third party, the behavior of one protein provides no additional information on the behavior of the other.}.

\added{While statistical associations and correlations are properties of the observed data, dependence and conditional independence are properties of the underlying processes that generate the data. Step 2 relies on statistical inference\cite{pearl2009causality,markowetz2007inferring} to infer from the observed data pairs of proteins that are conditionally independent. The conditionally independent pairs are ignored, and the remaining pairs are kept as hypothesized causal relations. This sparse inferred conditional independence graph is often desirable, as it reduces the many pairwise associations to a smaller number of hypothesized causal relations.}

Let's see how this applies to the MAPK signaling cascade. \added{Since Raf regulates the concentration of Erk by way of regulating Mek, Raf and Erk are dependent.  However, if we know the concentration of Mek, then the concentration of Raf provides no additional information about the concentration of Erk. Therefore, even though Raf and Erk are dependent, they are also conditionally independent given Mek.  }  Figure \ref{mapkInference}C illustrates the process of statistical inference by comparing the concentrations of Raf and Erk. When we subset the measurements to only the samples with high Mek, we can no longer see the association between Raf and Erk. \added{Formally, the algorithm tests the null hypothesis of conditional independence between Raf and Erk given the full range of values of Mek (and not just high values of Mek, as was shown for the purposes of illustration), and evaluates evidence against the null hypothesis\cite{spirtes2000causation}. In this example the test did not reject the null hypothesis, and resulted in removing the edge between Raf and Erk as in the middle graph of Figure \ref{mapk}A. }

At Step 2 the direction of the regulation remains unknown. Inference of the direction of the chain of events requires the experimental design, which involves external interventions or stresses. Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor. The intervention does not affect the concentration of Mek, however it blocks its ability to phosphorylate other proteins.  After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.  From this we can infer that Mek has causal influence on Erk. Since Raf was unaffected by the intervention, Mek does not have a causal influence on Raf, and therefore the direction of causal influence in this edge goes from Raf to Mek.  This intervention is required to transform the undirected graph in panel A - Step 2 to the causal graph in panel A - Step 3.

In the general case, computational methods for causal inference follow the workflow in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related proteins. Step 1 creates a dense network of pairwise associations. \added{Step 2 identifies cases of conditional independence, and removes the corresponding edges to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events. See Koller and Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings. Numerous implementations are available in statistical software, e.g. in the R package {\tt bnlearn} \cite{scutari2009learning}. \added{Depending on the biological system and on the experimental setting, the strength of causal evidence may vary. For example, Sachs, Itani {\it et al.} 2013\cite{sachs2013single} highlight conditions in phosho-proteomic experiments  where it may be infeasible to disentangle causality using perturbations.}

@@

\added{Experimentalists can choose to quantify more or less analytes, and collect more or less replicates such as proteins.} A typical high-throughput experiment includes \removed{a small number of interventions,} a small number of biological replicates, and quantifies a large number of analytes \removed{such as proteins}. \removed{This creates challenges in each of the 3 steps of causal inference above.}  \added{This experimental approach is an obstacle to causal inference.}

\added{To understand why, recall that we are interested in paring down the correlation graph in step one to the sparse conditional independence graph in step 2.  A {\it true positive} is the case when we correctly keep an edge from step 1 to step 2 in cases where the proteins are not conditionally independent.  Incorrectly keeping an edge when the proteins are actually conditionally independent is a case of a false positive.  Since the edges we preserve imply a regulatory relationship and motivate the expense of a validation experiment, false positives are costly and should be minimized.  However, the number of false positives increases as an exponential rate as one increases the number of proteins, much faster than the number of true positives.}

@@

We illustrate this problem with a computer simulation, inspired by Fan et. al \cite{fan2014challenges} but translated to our context.  We simulate data from a conditional independence network of 60 proteins and evaluate performance of structure inference algorithms in recovering this structure.  We then simulate a second experiment where the number of proteins is increased 99, to show how the increased protein coverage affects the performance of structure inference.  We varied depth as well as breadth -- In order to evaluate the effect of increasing sample size we start with 100 replicates, then repeat the two experiments with 200 replicates.  We evaluate performance by counting the total number of false positives (reject conditional independence for two conditionally independent proteins) and true positives (correct rejections).

The simplest case of for testing conditional independence is one of three variables -- two to compare, and at one variable upon which to condition the comparison.  Thus in the simulate we generate networks composed of three-protein subnetworks comprised of protein A, B, and C.  C's measurements come a univariate Gaussian distribution with 0 mean and variance 1.  Proteins A and B also have 0 mean and variance of 1, but their dependence with C is quantified in the multivariate Gaussian covariance matrix, each has a covariance with C of .3.  Most importantly, A and B are conditionally independent of each other given C.  So each subnetwork has two edges, from A to C and from B to C, and no edge from A to B is missing.  We created a network of 60 proteins within 20 disjoint subnetworks, 40 edges in total, and simulated multivariate Gaussian data with the covariance structure determined by this network.

We examined two cases for expanding the number of measured proteins.  The first describes the case of measuring proteins that provide no additional information about the biology.  We model this scenario by adding 39 unconnected proteins with independent Gaussian values for the second experiment.  Any inferred edge in the condition independence network linking these two proteins would be a false positive.  We then look at the case where more proteins are added, but the added proteins are informative about the biology.  We model this by increasing the number of triplets from 20 to 33, again to 99 proteins in total.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr2.png}}
\caption{Simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.  This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected. The histograms display the distributions of the maximum correlation values.Parts A and B indicates the case of randomly adding completely independent proteins to the panel, parts C and D illustrated informed addition of proteins -- the added proteins contain interdependencies.  The left column (A and C) show results for 30 replicates, the right (B and D) for 60 replicates. Comparing A and B to C and D demonstrate increasing the number of proteins on the panel creates more spurious correlation, and this is unaffected by whether or not the addition of proteins are added in an informative way.  This is a problem because correlation is used as initial evidence for a biological relationship.  Comparing the left and right columns show increasing the number of replicates can mitigates this somewhat.}
\label{spur_corr}
\end{figure}

Figure \ref{spur_corr} shows illustrates simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.  This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected. The histograms display the distributions of the maximum correlation values.  The top row (figure \ref{spur_corr} parts A and B) indicates the case if randomly adding proteins to the panel -- the added proteins where completely independent of one another protein in the network.  The bottom row (C and D) indicates the case of informed addition of proteins to the panel -- the added proteins contain interdependencies.  The first column (A and C) show results for 30 replicates, the second (B and D) for 60 replicates.  Parts A and B show that an increase in proteins leads to more high correlations, however since the added proteins were completely independent, the higher correlations were completely spurious.  Further, comparing parts C and D to A and B show that even when addition of proteins is informed by biology, the results are the same.  Spurious correlation increases with the number of proteins whether or not the addition of proteins is informed by biological prior knowledge.  This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.  Comparing the left and right columns show that increasing the number of replicates can mitigate this situation someone, but as we will see next, false positives will remain a problem.

This increased incidence of spurious correlations impedes Step 2, which \added{infers} conditional independences in the data.  To illustrate this, we repeated the previous simulation, but this time in each iteration we \added{performed Step 1 by calculating pairwise \added{Pearson} correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.}   \added{These algorithms have stricter data requirements that correlation calculations, so we changed settings to 60/100 samples, and 60/99 proteins to meet these requirements.}

Figure \ref{spur_dep} shows the counts of false positive edges in the resulting conditional independence graph, i.e. the number of protein pairs \added{where the null hypothesis of conditional independence was unduly rejected}.  The results demonstrate that an increase in the number of proteins leads to an increase in false positive rejections of conditional independence, and to a denser conditional independence graph.  \added{Again, comparing the top and bottom rows in figure \ref{spur_dep} show that it doesn't matter if the addition of proteins to the panel is informed or not. Increasing the number of proteins results in an exponential increase in pairwise comparisons -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of pairs.  So though the algorithm controls the rate of false positives, the absolute number of false positives increases exponentially.  Increasing the number of replicates does not help -- paradoxically in this case it hurt, most likely because the additional degrees of freedom enabled more tests, some of which added to the false positive count.} In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.  \added{ To verify that this pattern is not the artifact of a particular protein inference algorithm we repeated the simulation with two comparable algorithms\cite{tsamardinos2003algorithms,yaramakala2005speculative}, and arrived to the same conclusion (results not shown).}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep.png}}
\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples. The histograms display the counts of false positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation. The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.  The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Increasing the number of proteins results in more false positives without any biological justification.  Informed addition and added replicates does not help.}
\label{spur_dep}
\end{figure}

\added{These results lead to obvious question.  What exactly is the use of having more replicates? How does an informed selection of proteins for the panel make a difference relative to random selection?  These factors come to play in detection of true positives.  Figure\ref{true positives} is the true positive counterpart to figure\ref{true_pos}.  Given the described subnetwork approach we described above, the 60 protein network is comprised of 20 subnetworks of 3 proteins, each with 2 edges each for a total of 40 relationships that the algorithm is tasked with detecting.  That upper bound is illustrated with the dashed line in figure\ref{true_pos} parts A and B.  Uninformative addition of proteins adds no new relationships.  In parts C and D, informed addition of proteins adds 26 new relationships, indicated by the double-dashed line.  Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.  Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.  More importantly, the scale of the increase in total true positives gained from informed addition of proteins to the panel was less than that of the increase of false positives.  The costs of spurious correlation and resulting false positives quickly outweighs the benefits of higher throughput}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/true_pos.png}}
\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples. The histograms display the counts of true positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation. The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.  The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Dashed lines indicate the total amount of positives the algorithm is tasked with discovering.  Increasing the number of replicates increases true positives and is indeed essential when increasing the number of proteins increases the number of relationships to be detected.}
\label{spur_dep}
\end{figure}

\removed{A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes.  Once again the purpose of step 2 is to test conditional independence, and detect relationships between proteins that remain even when conditioning on other proteins.  While measuring too many analytes causes an explosion in false positives, collecting too few samples results in insufficient power to detect these relationships. While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 relies on having many replicates relative to the number analytes.  Note that even when the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high. The quality of causal inference approaches decreases drastically as the numbers of replicates fall.}

\added{Finally, the challenge of completing step 3 in the workflow also increases with the number of proteins quantified.}  Step 3 requires interventions to infer causal relationships from evidence of conditional independence.  \added{For a panel of k analytes, theoretical work by Eberhardt and Scheines {\it et al.} \cite{eberhardt2012number} show that in least ideal case of the only interventions available being interventions that target a single analyte (such as with small molecule inhibitors), k - 1 separate interventions would be needed.  With application of interventions that simultaneously target multiple analytes, this number can be lowered significantly, but this depends on several different factors\cite{eberhardt2012number,hauser2012two,hyttinen2013experiment} beyond the scope of this discussion.} \removed{For example, in the Raf-Mek-Erk pathway one intervention on Mek was sufficient to infer the two causal relationships. However, scaling this approach to a large number of proteins also introduces.}  But generally speaking as the number of proteins grows, so does the number of interventions required to fully infer causality, and eventually the total number of required interventions may become infeasible. Moreover, if the step 2 results in too many false positive hypothesized causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.

@@

\begin{enumerate}
\item \textit{Limit the number of analytes.} \added{Reducing the number of analytes dramatically reduces the number of spurious associations and resulting false positives.} Even though a list of analytes quantifiable with high-throughput technologies grows larger, only use a subset of measurements that are both biologically relevant and technologically accurate. The length of the list is not as important as the quality of measurements on the key parts of the system.  If the broader biological system is well understood, it may be possible to design a targeted experiment that focuses on a specific network or pathway, and ask more specific questions of the data, such as the presence of a particular regulatory event.  The more specific the question, the less data are needed to make solid causal conclusions.

\item \textit{Profile more biological replicates.}  \added{While reducing the number of analytes reduces spurious associations (false positives), increasing the number of replicates increases statistical power (true positives).}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population.  This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.  \added{A review by Twerve and Saez-Rodriguez provides a contrast of technologies by their content and sampling throughput\cite{SaezRodriguez:2012kx}.}

\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide much more data as an input to causal inference, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}. See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset. \cite{sachs2005causal}.
